\documentclass[11pt]{article}

\newcommand{\numpy}{{\tt numpy}}    % tt font for numpy

\topmargin -.5in
\textheight 9in
\oddsidemargin -.25in
\evensidemargin -.25in
\textwidth 7in

\begin{document}

% ========== Edit your name here
\author{Konstantin Krasser}
\title{DUA: }
\maketitle

\medskip

% ========== Begin answering questions here
Aufgabe 1 ( 9 points). In the lecture, you learned about the partition function. This function can not only be used to sort an array \(A\) (as in quicksort), but also to find the \(k\)-th smallest value in \(A\), i.e., the \(k\)-th element in the ascending sorted order of the elements in \(A\).
Example: In the array \(A=[6,7,2,9,3,1,0]\), the 4 th smallest element is the number 3 (only 0,1 , and 2 are smaller). We assume that every element in \(A\) is unique.
\begin{enumerate}

	\item Answer to question 1:

	      What would a naive approach, using comparison-based search algorithms, look like to find the \(k\)-th smallest element in \(A\) ? What is the lower asymptotic bound on the runtime of this approach?

	      The lower bound for any comparison-based algorithm is: The Naive approach would be to first sort the array and then return the \(k-th\) element. Thus making the lower bound for:
	      \begin{equation}
		      (O) n \log n
	      \end{equation}

	\item Answert to question 2:

	      (2 points) Provide a modified partition function in pseudocode that rearranges the array \(A\) such that the pivot element is at position \(i_p\), and all elements to the left of \(i_p\) are smaller than the pivot element, and all elements to the right of \(i_p\) are greater than the pivot element. The manipulation of \(A\) is done in-place, so all changes are made directly in \(A\), and partition does not need to explicitly return the array \(A\), only \(i_p\).


	\item Answer to question 3:

	      (4 points) Describe in words and in pseudocode how the modified partition function can be used to efficiently find the \(k\)-th smallest value in \(A\).


	\item Answer to question 4:

	      (2 points) What is the runtime of your algorithm in the best case and in the worst case? Provide an example call for both cases. The best/worst case should apply to general \(k\), not a specific \(k\).

\end{enumerate}
\end{document}
